LabMate: a prospectus for types for MATLAB

### 2024-04-11 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] picture of a matrix

\newcommand{\todo}[1]{\textcolor{red}{\textbf{TODO: #1}}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Under the hood, LabMate translates MATLAB expressions to typed
expressions in its own core type theory.
%
Since matrices feature heavily in MATLAB code, the type of matrices
plays a central role in this type theory.
%
We make this type rather precise, by allowing the type of each entry
in the matrix to vary according to the row and column of the entry:
%
we parameterise the type of matrices by a type $R$ associated with
each row, a type $C$ associated with each column, and a type $E(x,y)$
which might vary depending on the values of parameters $x : R$ and
$y : C$, as well as two lists $\rs : \List{R}$ and $\List{C}$ --- the
idea being that the $i$th entry in the list $\rs$ and the $j$th entry
in the list $\cs$ determines the type of the $(i, j)$ coordinate in the
matrix.
%
Matrix types formation can thus be summarised as the inference rule
given in Figure~\ref{fig:matrix_intro}.
%\[
%  \infer{\istype{\Matrix{R}{C}{E}{rs}{\cs}}}
%  {
%    \istype{R}
%    &\quad
%    \istype{C}
%    &\quad
%    x : R, y : C \vdash \istype{E(x,y)}
%    &\quad
%    rs : \List{R}
%    &\quad
%    cs : \List{C}
%  }
%\]
%
\begin{figure*}[th]
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{\istype{\Matrix{R}{C}{E}{\rs}{\cs}}}
      {
        \istype{R}
        &\quad
        \istype{C}
        &\quad
        x : R, y : C \vdash \istype{E(x,y)}
        &\quad
        \rs : \List{R}
        &\quad
        \cs : \List{C}
      }
    \]
  \end{center}
  \caption{Type formation rule for matrix types}
  \label{fig:matrix_intro}
\end{subfigure}
\hfill
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{A + B : \Matrix{R}{C}{E}{\rs}{\cs}}
      {
        x : R, y : C \vdash \isadd{E(x, y)}
        &\quad
        A : \Matrix{R}{C}{E}{\rs}{\cs}
        &\quad
        B : \Matrix{R}{C}{E}{\rs}{\cs}
      }
    \]
  \end{center}
  \caption{Addition rule for matrices}
  \label{fig:matrix_add}
\end{subfigure}
\hfill
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{A * B : \Matrix{R}{C}{E''}{\rs}{\cs}}
      {
        \ismult{E}{E'}{E''}
        &\quad
        A : \Matrix{R}{C}{E}{\rs}{\ms}
        &\quad
        B : \Matrix{R}{C}{E'}{\ms}{\cs}
      }
    \]
  \end{center}
  \caption{Multiplication rule for matrices}
  \label{fig:matrix_mul}
\end{subfigure}
% FIXME : \caption is broken
\captionwide{Typing rules for matrices}
\label{fig:matrix_rules}
\end{figure*}

>>>>

\newcommand{\todo}[1]{\textcolor{red}{\textbf{TODO: #1}}}

\newcommand{\param}{parametrize}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Under the hood, LabMate translates selected well behaved MATLAB expressions
to typed expressions in its own core theory of types and values.
%
As matrices feature heavily in MATLAB code, the types of matrices
play a central role in this theory.
%
In particular, we acknowledge that the rows and columns of a matrix might pertain to distinct individual entities, rather as the rows ans columns of a spreadsheet do.
%
A matrix type looks like
\[\Matrix{R}{C}{E}{\rs}{\cs}
\]
%
We thus \param{} matrix types by two arbitrary types $R$, for the individuals the rows concern, and $C$ for the columns. We may then give two lists, $\rs = [r_1,\ldots,r_m]\in\List{R}$ and $\cs = [c_1,\ldots,c_n]\in\List{C}$ of those individuals. A \param{}d type $E(r,c)$ gives the type of entry for any combination of row and column individual, with $E(r_i,c_j)$ being the type of the entry with coordinates $(i,j)$.
\[\begin{array}{r|ccccc|}
\;     & c_1      & \cdots & c_j     & \cdots & \multicolumn{1}{c}{c_n}     \\
\hline
r_1    & e_{1,1}  &        & e_{1,j} &        & e_{1,n} \\
\vdots &          &        &         &        &         \\
r_i    & e_{i,1}  &        & e_{i,j} &        & e_{i,n} \\
\vdots &          &        &         &        &         \\
r_m    & e_{m,1}  &        & e_{m,j} &        & e_{m,n}  \\
      \cline{2-6}
\end{array}\]

We make this type rather precise, by allowing the type of each entry
in the matrix to vary according to the row and column of the entry:
%
we \param{} the type of matrices by a type $R$ associated with
each row, a type $C$ associated with each column, and a type $E(x,y)$
which might vary depending on the values of parameters $x : R$ and
$y : C$, as well as two lists $\rs : \List{R}$ and $\List{C}$ --- the
idea being that the $i$th entry in the list $\rs$ and the $j$th entry
in the list $\cs$ determines the type of the $(i, j)$ coordinate in the
matrix.
%
Matrix types formation can thus be summarised as the inference rule
given in Figure~\ref{fig:matrix_intro}.
%\[
%  \infer{\istype{\Matrix{R}{C}{E}{rs}{\cs}}}
%  {
%    \istype{R}
%    &\quad
%    \istype{C}
%    &\quad
%    x : R, y : C \vdash \istype{E(x,y)}
%    &\quad
%    rs : \List{R}
%    &\quad
%    cs : \List{C}
%  }
%\]
%
\begin{figure*}[th]
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{\istype{\Matrix{R}{C}{E}{\rs}{\cs}}}
      {
        \istype{R}
        &\quad
        \istype{C}
        &\quad
        x : R, y : C \vdash \istype{E(x,y)}
        &\quad
        \rs : \List{R}
        &\quad
        \cs : \List{C}
      }
    \]
  \end{center}
  \caption{Type formation rule for matrix types}
  \label{fig:matrix_intro}
\end{subfigure}
\hfill
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{A + B : \Matrix{R}{C}{E}{\rs}{\cs}}
      {
        x : R, y : C \vdash \isadd{E(x, y)}
        &\quad
        A : \Matrix{R}{C}{E}{\rs}{\cs}
        &\quad
        B : \Matrix{R}{C}{E}{\rs}{\cs}
      }
    \]
  \end{center}
  \caption{Addition rule for matrices}
  \label{fig:matrix_add}
\end{subfigure}
\hfill
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{A * B : \Matrix{R}{C}{E''}{\rs}{\cs}}
      {
        \ismult{E}{E'}{E''}
        &\quad
        A : \Matrix{R}{C}{E}{\rs}{\ms}
        &\quad
        B : \Matrix{R}{C}{E'}{\ms}{\cs}
      }
    \]
  \end{center}
  \caption{Multiplication rule for matrices}
  \label{fig:matrix_mul}
\end{subfigure}
% FIXME : \caption is broken
\captionwide{Typing rules for matrices}
\label{fig:matrix_rules}
\end{figure*}
